A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-04-16 submission/manuscript.rmd
fix titles
include references

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
editor_options: 
  chunk_output_type: console
csl: jamia.csl
---

@@

# Introduction
Interest in understanding how a treatment’s effect varies across patients—a
concept described as heterogeneity of treatment effects (HTE)—has been growing.
This concept is central to the agenda for both personalized (or precision)
medicine and comparative effectiveness research. More formally, HTE has been
defined as non-random variability in the direction or magnitude of a treatment
effect, in which the effect is measured using clinical outcomes. Usually,
analyses focus on the relative scale, where treatment effects are assessed one
at a time in patient subgroups defined from single covariates, an approach that
suffers from low power and multiplicity issues. However, even with
well-established constant relative effects, treatment benefit (or harm) may vary
substantially on the absolute scale.

More recently, “predictive” HTE analyses have been described (and contrasted
with “one-variable-at-a-time” subgroup analysis) as approaches that provide
predictions of potential outcomes in a particular patient with one intervention
versus an alternative, taking into account multiple relevant patient
characteristics. One promising approach is “risk modeling”, in which treatment
effects are estimated  in strata of predicted risk. Such a risk-based approach
first stratifies patients according to baseline risk predictions, using either
an existing or an internally developed risk prediction model. Then, relative and
absolute treatment effects are estimated within risk strata.

While these approaches have generally been recommended for application to
clinical trials, observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions. They are also often larger than many
typical trials, providing excellent power for HTE analysis, and include
heterogeneous populations. However, unlike trials, treatment effects are subject
to confounding and the unique structure of different databases calls for
database-specific analysis plans that are often not easily transportable.

The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
established an international network of data partners and researchers that aim
to bring out the value of health data through large-scale analytics by mapping
all available databases to the Observational Medical Outcomes Partnership (OMOP)
Common Data Model (CDM). The common data structure enables analyses at a very
large scale. For example, in a recent study, a large set of first-line
treatments for hypertension was compared with respect to 55 outcomes in a
network of databases, including 4.9 million patients from around the world.

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in high-dimensional observational data. We implemented the
framework using existing OHDSI methods for use in the OMOP-CDM, including the
patient-level prediction framework and the population-level effect estimation
framework based on new-user cohort design. As a proof-of-concept we analyzed
heterogeneity of the effects of first-line hypertension treatment: we compared
the effect of angiotensin converting enzyme (ACE) inhibitors to beta blockers on
9 outcomes across three different US claims databases.

@@

## Step 3: Prediction We adopt the standardized framework for the generation of
patient-level prediction models using observational data that ensures adherence
to existing guidelines. This prediction framework requires the definition of two
essential cohorts: a target cohort and an outcome cohort.

To generate the target cohort we pool the already defined treatment cohort  and
comparator cohort . However, for risk-based analysis of treatment effects it is
necessary to avoid deferentially fitting the prediction model to patients across
the treatment arm in order to avoid inducing spurious interactions. To do this,
we developed the patient-level prediction model in the propensity score-matched
subset of the population (1:1), where treatment assignment is well-balanced. The
propensity scores are based on LASSO logistic regression for modeling the
association between treatment assignment and all available demographics, drug
exposures, diagnoses, measurements and medical procedures. Finally, we need to
define the time horizon for which we aim to make predictions and we need to
select the machine-learning algorithm we want to use to generate patient-level
predictions. Currently, the available options are regularized logistic
regression, random forest, gradient boosting machines, decision tree, naive
Bayes, K-nearest neighbors, neural network and deep learning (convolutional
neural networks, recurrent neural network and deep nets).

@@

## Step 5: Result presentation and evaluation Our framework provides
standardized output for each step of the analysis. The number of patients and
person years by treatment arm along with the number of outcomes. A performance
overview of the derived prediction models, including discrimination and
calibration both in the propensity score matched subset, the entire population
and separately for treated and comparator patients. This is rather relevant as
the performance of the prediction models is directly related to our ability to
single out patient subgroups where treatment may be highly beneficial or unsafe.
Kent et al  demonstrated that the event rate and the discriminative ability of
the prediction model can predict very well the distribution of predicted risk.
The lower the event rate and the higher the c-statistic (given good calibration)
result in high risk heterogeneity, thus making estimated average treatment
effects uninformative. In this case, risk stratified analysis of HTE can be more
effective in singling out patient subgroups that stand to benefit (or be harmed)
most by treatment in question.

@@

# Results
As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
common treatment classes for hypertension, with a well-established
effectiveness. Beta blockers, even though initially widely used for the
treatment of hypertension, more recent trials and meta-analyses have cast doubt
on their relative effectiveness. As a result, newer US guidelines do not
consider them for initial treatment for hypertension while in the EU guidelines
combination with other antihypertensive treatments is recommended. However,
another meta-analysis suggested that the efficacy profile of beta blockers is
similar to other major treatment classes in younger hypertensive patients and,
thus, countries like Canada still include them as a first-line candidate for the
treatment.

@@

This example nicely illustrates heterogeneity of absolute treatment effects,
i.e. differences in absolute benefits and harms of ACE-inhibitors vs beta
blockers for patients with different baseline risk. The results suggest that
treatment with ACE-inhibitors, compared to treatment with beta blockers, may be
focused on the higher risk patients, in whom the benefits outweigh the harms.
However, treatment with beta blockers may be a viable option in lower risk
patients, in whom the benefit-harm tradeoff is in favor of beta blockers. This
is in accordance with earlier findings that beta blockers should be considered
as first-line treatment for younger hypertensive patients. More thorough
evaluation of these results is required in future research.

@@

# Discussion
We developed a framework for the assessment of heterogeneity of treatment effect
in large observational databases using a risk modeling approach. The framework
is implemented in an open source R-package in the OHDSI methods library
(https://github.com/OHDSI/RiskStratifiedEstimation). As a proof-of-concept, we
used our framework to evaluate heterogeneity of the effect of treatment with
ACE-inhibitors compared to beta blockers on 3 efficacy and 6 safety outcomes.

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs. However, low power and
restricted prior knowledge on the mechanisms of variation in treatment effect
are often inherent in RCTs, which are often adequately powered only for the
analysis of the primary outcome. Observational databases contain a large amount
of information on treatment assignment and outcomes of interest, while also
capturing key patient characteristics. Our framework provides a standardized
approach that can be used to leverage available information from these data
sources, allowing for large-scale assessment of treatment effect heterogeneity.
Multiple outcomes can be evaluated in patient subgroups of similar baseline
outcome risk, where different outcome risk stratification schemes can be
considered. The standardized nature of the framework enables transportability to
multiple databases, provided that they are mapped to the OMOP-CDM.

Recently, guidelines on the application of risk modeling approaches for the
assessment of heterogeneity of treatment effect in RCT settings have been
proposed. Our framework aims to translate these guidelines to the observational
setting while also providing a toolset for its implementation. Several
considerations need to be made. First, estimates may be biased due to the
observational nature of the data. We attempt to account for potential
confounding by estimating propensity scores within strata of predicted risk.
These scores are estimated using regularized logistic regression on a large set
of pre-defined covariates. This specific approach gave accurate results in
extensive simulation studies. However, such approaches do not account for
unobserved confounding. Several sensitivity analyses have been proposed in the
literature for measuring the robustness of results in the presence of unobserved
confounding. Another approach is to calibrate estimates and confidence intervals
based on a large set of negative controls. Negative controls are
treatment-outcome pairs for which a null effect has been established. Estimating
these effects within available data provides an approximation of the null
distribution that can be used to empirically recalibrate effect estimates.
Future work may extend our framework with this type of analyses.

Our method provides a risk-stratified assessment of treatment effect
heterogeneity. However, even though stratification can provide a rough guide for
clinical interpretation, it is not appropriate to guide clinical practice, where
decisions need to be made at the individual level. Presentation of treatment
effects as a continuous function of risk would be more helpful, but is
methodologically challenging. Future research is necessary for the development
of methods for continuous risk-based assessment of HTE.

Externally derived prediction models are preferred for analyzing treatment
effect heterogeneity. Such external models should be well transportable. In the
absence of such prediction models, simulations of RCTs have shown that
internally derived models can be used to provide unbiased estimates of treatment
effect across the spectrum of baseline risk. However, in observational databases
treatment arms may significantly differ in sample size. Because the prediction
model will possibly better fit to the larger treatment arm, this may introduce
spurious treatment-covariate interactions in the prediction model, leading to
sub-optimal risk stratification. As a remedy, we first match the patients in the
treatment and the comparator cohorts on the basis of propensity scores.
Additionally, we propose to assess model performance in the separate treatment
arms to evaluate its aptness for risk stratification.

@@

Recently, disease risk scores have been explored as an alternative to propensity
scores for balancing covariates. In our method, the objective of risk
stratification is not balancing, but assessing the variation of treatment
effects on multiple outcomes across patients with different levels of baseline
risk. Although using the same risk model for balancing and risk-based HTE
analysis may sound attractive, we note that our method only uses one risk model
for stratification and one propensity score model for balancing, while separate
disease risk score models would be required to analyze treatment effects for
each of the multiple outcomes.

@@

>>>>

---
title: "A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left, pagewise]{lineno}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jamia.csl
---

@@

# Introduction
Interest in understanding how a treatment’s effect varies across patients—a
concept described as heterogeneity of treatment effects (HTE)—has been growing.
This concept is central to the agenda for both personalized (or precision)
medicine and comparative effectiveness research. More formally, HTE has been
defined as non-random variability in the direction or magnitude of a treatment
effect, in which the effect is measured using clinical outcomes
[@KRAVITZ2004]. Usually, analyses focus on the relative scale, where treatment
effects are assessed one at a time in patient subgroups defined from single
covariates, an approach that suffers from low power and multiplicity issues
[@Varadhan2013]. However, even with well-established constant relative
effects, treatment benefit (or harm) may vary substantially on the absolute
scale.

More recently, “predictive” HTE analyses have been described (and contrasted
with “one-variable-at-a-time” subgroup analysis) as approaches that provide
predictions of potential outcomes in a particular patient with one intervention
versus an alternative, taking into account multiple relevant patient
characteristics. One promising approach is “risk modeling”, in which treatment
effects are estimated in strata of predicted risk [@Kent2010; @Kent2018]. Such
a risk-based approach first stratifies patients according to baseline risk
predictions, using either an existing or an internally developed risk prediction
model [@Kent2010]. Then, relative and absolute treatment effects are estimated
within risk strata.

While these approaches have generally been recommended for application to
clinical trials, observational databases are also an appealing substrate.
Observational healthcare databases, such as administrative claims and electronic
health records, are already highly available for the analysis of
pharmacoepidemiologic research questions [@AdlerMilstein2017; @Dahabreh2014].
They are also often larger than many typical trials, providing excellent power
for HTE analysis, and include heterogeneous populations. However, unlike trials,
treatment effects are subject to confounding and the unique structure of
different databases calls for database-specific analysis plans that are often
not easily transportable.

The Observational Health Data Sciences and Informatics (OHDSI) collaborative
(!!REF) has established an international network of data partners and
researchers that aim to bring out the value of health data through large-scale
analytics by mapping all available databases to the Observational Medical
Outcomes Partnership (OMOP) Common Data Model (CDM) [@Overhage2012]. The
common data structure enables analyses at a very large scale. For example, in a
recent study [@Suchard2019], a large set of first-line treatments for
hypertension was compared with respect to 55 outcomes in a network of databases,
including 4.9 million patients from around the world.

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in high-dimensional observational data. We implemented the
framework using existing OHDSI methods for use in the OMOP-CDM, including the
patient-level prediction framework [@Reps2018] and the population-level effect
estimation framework [@Ryan2013] based on new-user cohort design. As a
proof-of-concept we analyzed heterogeneity of the effects of first-line
hypertension treatment: we compared the effect of angiotensin converting enzyme
(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
databases.

@@

## Step 3: Prediction 
We adopt the standardized framework for the generation of patient-level
prediction models using observational data that ensures adherence to existing
guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
definition of two essential cohorts: a target cohort and an outcome cohort.

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort . However, for risk-based analysis of treatment effects it is
necessary to avoid deferentially fitting the prediction model to patients across
the treatment arm in order to avoid inducing spurious interactions [@Burke2014;
@vanKlaveren2019]. To do this, we developed the patient-level prediction model
in the propensity score-matched subset of the population (1:1), where treatment
assignment is well-balanced. The propensity scores are based on LASSO logistic
regression for modeling the association between treatment assignment and all
available demographics, drug exposures, diagnoses, measurements and medical
procedures. Finally, we need to define the time horizon for which we aim to make
predictions and we need to select the machine-learning algorithm we want to use
to generate patient-level predictions. Currently, the available options are
regularized logistic regression, random forest, gradient boosting machines,
decision tree, naive Bayes, K-nearest neighbors, neural network and deep
learning (convolutional neural networks, recurrent neural network and deep
nets).

@@

## Step 5: Result presentation and evaluation 
Our framework provides standardized output for each step of the analysis. The
number of patients and person years by treatment arm along with the number of
outcomes. A performance overview of the derived prediction models, including
discrimination and calibration both in the propensity score matched subset, the
entire population and separately for treated and comparator patients. This is
rather relevant as the performance of the prediction models is directly related
to our ability to single out patient subgroups where treatment may be highly
beneficial or unsafe.  Kent et al [@Kent2016] demonstrated that the event rate and the
discriminative ability of the prediction model can predict very well the
distribution of predicted risk. The lower the event rate and the higher the
c-statistic (given good calibration) result in high risk heterogeneity, thus
making estimated average treatment effects uninformative. In this case, risk
stratified analysis of HTE can be more
effective in singling out patient subgroups that stand to benefit (or be harmed)
most by treatment in question.

@@

# Results
As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
common treatment classes for hypertension, with a well-established
effectiveness. Beta blockers, even though initially widely used for the
treatment of hypertension, more recent trials and meta-analyses have cast doubt
on their relative effectiveness (!!REF). As a result, newer US guidelines do
not consider them for initial treatment for hypertension while in the EU
guidelines combination with other antihypertensive treatments is recommended
[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
the efficacy profile of beta blockers is similar to other major treatment
classes in younger hypertensive patients and, thus, countries like Canada still
include them as a first-line candidate for the treatment [@Khan2006;
@Rabi2020].

@@

This example nicely illustrates heterogeneity of absolute treatment effects,
i.e. differences in absolute benefits and harms of ACE-inhibitors vs beta
blockers for patients with different baseline risk. The results suggest that
treatment with ACE-inhibitors, compared to treatment with beta blockers, may be
focused on the higher risk patients, in whom the benefits outweigh the harms.
However, treatment with beta blockers may be a viable option in lower risk
patients, in whom the benefit-harm tradeoff is in favor of beta blockers. This
is in accordance with earlier findings that beta blockers should be considered
as first-line treatment for younger hypertensive patients [@Cruickshank2007; @Khan2006]. More thorough
evaluation of these results is required in future research.

@@

# Discussion
We developed a framework for the assessment of heterogeneity of treatment effect
in large observational databases using a risk modeling approach. The framework
is implemented in an open source R-package in the OHDSI methods library
(<https://github.com/OHDSI/RiskStratifiedEstimation>). As a proof-of-concept, we
used our framework to evaluate heterogeneity of the effect of treatment with
ACE-inhibitors compared to beta blockers on 3 efficacy and 6 safety outcomes.

In recent years several methods for the evaluation of treatment effect
heterogeneity have been developed in the setting of RCTs [@Rekkas2020].
However, low power and restricted prior knowledge on the mechanisms of variation
in treatment effect are often inherent in RCTs, which are often adequately
powered only for the analysis of the primary outcome. Observational databases
contain a large amount of information on treatment assignment and outcomes of
interest, while also capturing key patient characteristics. Our framework
provides a standardized approach that can be used to leverage available
information from these data sources, allowing for large-scale assessment of
treatment effect heterogeneity. Multiple outcomes can be evaluated in patient
subgroups of similar baseline outcome risk, where different outcome risk
stratification schemes can be considered. The standardized nature of the
framework enables transportability to multiple databases, provided that they are
mapped to the OMOP-CDM.

Recently, guidelines on the application of risk modeling approaches for the
assessment of heterogeneity of treatment effect in RCT settings have been
proposed [@Kent2019; @PathEnE]. Our framework aims to translate these
guidelines to the observational setting while also providing a toolset for its
implementation. Several considerations need to be made. First, estimates may be
biased due to the observational nature of the data. We attempt to account for
potential confounding by estimating propensity scores within strata of predicted
risk. These scores are estimated using regularized logistic regression on a
large set of pre-defined covariates. This specific approach gave accurate
results in extensive simulation studies [@Tian2018]. However, such approaches
do not account for unobserved confounding [@Liu2013]. Several sensitivity
analyses have been proposed in the literature for measuring the robustness of
results in the presence of unobserved confounding. Another approach is to
calibrate estimates and confidence intervals based on a large set of negative
controls [@Schuemie2014; @Schuemie2571]. Negative controls are
treatment-outcome pairs for which a null effect has been established. Estimating
these effects within available data provides an approximation of the null
distribution that can be used to empirically recalibrate effect estimates.
Future work may extend our framework with this type of analyses.

Our method provides a risk-stratified assessment of treatment effect
heterogeneity. However, even though stratification can provide a rough guide for
clinical interpretation, it is not appropriate to guide clinical practice, where
decisions need to be made at the individual level [@Kent2019]. Presentation of
treatment effects as a continuous function of risk would be more helpful, but is
methodologically challenging. Future research is necessary for the development
of methods for continuous risk-based assessment of HTE.

Externally derived prediction models are preferred for analyzing treatment
effect heterogeneity [@Kent2010]. Such external models should be well
transportable. In the absence of such prediction models, simulations of RCTs
have shown that internally derived models can be used to provide unbiased
estimates of treatment effect across the spectrum of baseline risk
[@Burke2014]. However, in observational databases treatment arms may
significantly differ in sample size. Because the prediction model will possibly
better fit to the larger treatment arm, this may introduce spurious
treatment-covariate interactions in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.

@@

Recently, disease risk scores have been explored as an alternative to propensity
scores for balancing covariates [@Glynn2012; @Hansen2008]. In our method, the
objective of risk stratification is not balancing, but assessing the variation
of treatment effects on multiple outcomes across patients with different levels
of baseline risk. Although using the same risk model for balancing and
risk-based HTE analysis may sound attractive, we note that our method only uses
one risk model for stratification and one propensity score model for balancing,
while separate disease risk score models would be required to analyze treatment
effects for each of the multiple outcomes.

@@

\newpage
# References
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent
<div id="refs"></div>
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent
